Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2025-03-23 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\usepackage[labelfont=bf]{caption}
% \usepackage{showframe}
% \usepackage{layout}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{loextfig}}{\newfloat{suppfig}{h}{loextfig}[chapter]}
\floatname{suppfig}{Extended Data Fig.}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Extended Data Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the ordinary differential equation (ODEs) proposed in
velocyto \citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
\end{align} where \(u(t), s(t)\) are the unspliced and spliced
expression levels of a gene at time \(t\) under a transcription rate
\(\alpha(t)\) with possible temporal dependence, splicing rate
\(\beta\), and degradation rate \(\gamma\). We specify this model to a
setting that depends on cell \(c\) and gene \(g\) as follows:
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
\end{align} In the equation, the subscript \(c\) is the cell dimension,
\(g\) is the gene dimension, \(\left( u_{cg}(t), s_{cg}(t) \right)\) are
the unspliced and spliced expression functions given the change of time
per cell and gene. We restrict attention to piecewise-constant
\(\alpha_{cg}(t)\) to capture gene-specific activation and repression.
We take special care to model a gene- and cell-specific switching time
that marks a single transition from activation to repression by
introducing a Bernoulli variable \(k_{cg}\) to model unknown activation
state. We assume our cell-by-gene data-matrix arrive as observations of
Poisson-counts related to the solution of the above ODEs at unknown
times \(\tau_{cg}\), which is modeled as a relationship between an
unknown latent time shared across each cell, \(t_c\), and unknown
gene-specific time-offsets \(t_{0,g}\) where all read counts for a
single cell occurred at an unknown, but shared latent time \(t_c\).
These relative times are also used to parametrize the Bernoulli process
for \(k_{cg}\). Importantly, we recognize that the initial conditions
are in fact unknown.

@@

Next we introduce the priors for the splicing parameters (where the
activation rate \(\alpha\) depends on the activation state \(k_{cg}\)
from above): \begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1
  \end{cases}
\end{align}
\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}

@@

We define the ODE solution at time \(\tau_{cg}\) as: \begin{equation}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

Next, we define the observation model that gives rise to the observed
counts as: \begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align} Here, we use
\({u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}\) to denote the
observed unspliced and spliced counts for cell \(c\) and gene \(g\). We
use \(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}\) to
denote our generative model's prediction of these unspliced and spliced
expression levels. The generative process for modeling these observed
read counts given denoised gene transcript expression level
\(\hat{u}_{cg}, \hat{s}_{cg}\) considers the expected number of observed
reads for a given gene in a given cell as the number of transcripts
times the ratio of the cell's total reads to total transcripts.
\textbf{Improve descriptions of how noise is modeled in the observation model.}

\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
The analytic solution of the differential equations to predict spliced
and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
\citep{Bergen2020-pj, Li2021-qa} and given in Eqs.
\ref{eq-solution-u}-\ref{eq-solution-s2}. \begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

To simplify these equations, consider the case when \(k_{cg} = 0\) and
\(\beta_g \neq \gamma_g\). Then, \begin{align}
u\left(\tau^{(0)}\right) &= {u_0^{(0)}}_g e^{-\beta_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt + \frac{{\alpha^{(0)}}_g}{\beta_g}\left(1-e^{-\beta_g \tau^{(0)}}\right), 
  \label{eq-sol-usimp} \\
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}\left(1-e^{-\gamma_g \tau^{(0)}}\right) 
  \nonumber\\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(0)}}-e^{-\beta_g \tau^{(0)}}\right). \label{eq-sol-ssimp}
\end{align} When \(k_{cg} = 0\) and \(\beta_g = \gamma_g\), then
\(u\left(\tau^{(0)}\right)\) has the same solution, and
\(s\left(\tau^{(0)}\right)\) becomes \begin{align}
s\left(\tau^{(0)}\right) &= s_0^{(0)}{ }_g e^{-\gamma_g \tau^{(0)}} \nonumber \\
  & \hskip -24pt +\frac{{\alpha^{(0)}}_g}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{(0)}}\right) \nonumber\\
  & \hskip -24pt - \left( {\alpha^{(0)}}_g-\beta_g {u_0^{(0)}}_g \right) 
    \tau^{(0)} e^{-\beta_g \tau^{(0)}}. \label{eq-sol-ssimp2}
\end{align} When \(k_{cg} = 1\) and \(\beta_g \neq \gamma_g\), then
\begin{align}
u\left(\tau^{(1)}\right) &=u_0^{(1)} g^{e^{-\beta_g \tau^{(1)}}}, \\
s\left(\tau^{(1)}\right) &=s_0^{(1)} e^{-\gamma_g \tau^{(1)}} \nonumber \\
  & \hskip -24pt +\frac{-\beta_g u_0^{(1)}}{\gamma_g-\beta_g}
  \left(e^{-\gamma_g \tau^{(1)}}-e^{-\beta_g \tau^{(1)}}\right).
\end{align} When \(k_{cg} = 1\) and \(\beta_g = \gamma_g\), then
\(u\left(\tau^{(1)}\right)\) has the same solution, and
\(s\left(\tau^{(1)}\right)\) becomes \begin{align}
s\left(\tau^{(1)}\right)=s_0^{(1)}{ }_g e^{-\gamma_g \tau^{(1)}}
  +\beta_g u_0^{(1)}{ }_g \tau^{(1)} e^{-\beta_g \tau^{(1)}}.
\end{align}

@@

Given observations
\(\tilde{X}_{cg} = \left( u_{cg}^{obs}, s_{cg}^{obs} \right)\), we would
like to compute the posterior distribution over the random variables
\begin{align}
\theta &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi &= \left( {t_{0}^{(0)}}_g, \Delta \text{switching}_{g}, 
          {\alpha^{(0)}}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align} but exact Bayesian inference is intractable in this model.
We use Pyro to automatically integrate out the local discrete latent
variables \(k\), which is defined as the cell and gene transcriptional
state (see above), and approximate the posterior over the remaining
latent variables using variational inference
\citep{Bingham2018-id, Kucukelbir2016-bk}, which converts intractable
integrals into optimization problems that can be solved with
off-the-shelf tools. In variational inference, we maximize a tractable
bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
divergence \[
\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
\] between a parametric family of tractable probability distributions q
and the intractable true posterior \begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align} We approximate our model's posterior distribution
\(p( \theta, \psi \vert
\tilde{X})\) with a tractable family of probability distributions
\(q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)\), where
\(q_{\phi_1}(\theta)\) is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
\(q_{\phi_2}(\psi)\) is a multivariate Gaussian distribution with a
learnable location parameter and a learnable low-rank covariance matrix.
We solve the resulting stochastic optimization problem using a version
of Automatic Differentiation Variational Inference (ADVI)
\citep{Kucukelbir2016-bk}: we obtain gradient estimates by
differentiating a Monte Carlo estimate of the ELBO and update the
variational parameters using stochastic gradient ascent. Our
implementation and experiments use generic variational families and
Monte Carlo ELBO and gradient estimators provided by Pyro
\citep{Bingham2018-id}, an open source library for probabilistic machine
learning, and a version built into Pyro of the adaptive stochastic
gradient ascent algorithm Adam augmented with gradient clipping to
enhance numerical stability during training.

@@

To benchmark Pyro -Velocity performance in predicting cell fate, we
generated the posterior samples measurement
\(x_{cg}=\left(u\left(\tau^{(k_{cg})}\right),
s\left(\tau^{(k_{cg})}\right)\right)\) or
\(x_{cg}=\left(u_{cg}, s_{cg} \right)\), \(t_c\), \(\beta_g\),
\(\gamma_g\) from the same single cell RNA-seq using \(N=30\) Monte
Carlo samples from the posterior predictive distribution following
\(p(x_{cg} \vert \theta, \psi) p(\theta, \psi) \approx p(x_{cg} \vert \theta,
\psi) q_{\phi}(\theta, \psi)\). \(x_{cg}\) can be posterior samples of
either denoised gene expression (used for phase portraits and vector
field-based trajectory inference) or raw read counts (used for gene
selection). Then we calculated the posterior samples of RNA velocity as
\(v_{cg}=\frac{ds(\tau^{(k_{cg})})}{d \tau^{(k{cg})}}\) based on
posterior samples measurement of denoised gene expression \(x_{cg}\) and
\(\beta_{g}\), \(\gamma_{g}\).

@@

We prioritize the cell fate markers using two metrics: First, the
Pearson correlation between each gene's posterior mean of the denoised
spliced expression and posterior mean of the shared time; Second, the
negative mean absolute errors of each gene's observed spliced, and
unspliced read counts with posterior predictive samples of spliced and
unspliced raw read counts, i.e.
\(\frac{1}{N n_c} \sum_i \sum_c (x_{icg}-\tilde{X}_{cg})\), where \(N\)
is the posterior sample number that is set to \(30\), \(i\) is the
posterior sample index, and \(n_c\) is the cell number. We first select
the top \(300\) genes with the highest negative mean absolute errors and
then rank the \(300\) genes based on the most positively correlated
genes and the least negatively correlated genes. We use the same
strategy for scVelo to rank the markers by the model likelihood to get
the top 300 genes and then prioritize these genes by Pearson's
correlation between scVelo latent time and normalized expression.

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

@@

\raggedbottom
\usepackage{nameref}
\usepackage{placeins}
\usepackage[utf8]{inputenc}
\usepackage{fontspec}
\usepackage[labelfont=bf]{caption}
% \usepackage{showframe}
% \usepackage{layout}
\allowdisplaybreaks
\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{loextfig}}{\newfloat{suppfig}{h}{loextfig}[chapter]}
\floatname{suppfig}{Extended Data Fig.}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Extended Data Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

@@

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the ordinary differential equation (ODEs) proposed in
velocyto \citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}
\begin{equation}
  \label{eq-rna}
\begin{aligned}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0,
\end{aligned}
\end{equation} where \(u(t), s(t)\) are the unspliced and spliced
expression levels of a gene at time \(t\) under a transcription rate
\(\alpha(t)\) with possible temporal dependence, splicing rate
\(\beta\), and degradation rate \(\gamma\). We specify this model to a
setting that depends on cell \(c\) and gene \(g\) as follows:
\begin{equation}
    \label{eq-rna-cg}
\begin{aligned}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} .
\end{aligned}
\end{equation} In the equation, the subscript \(c\) is the cell
dimension, \(g\) is the gene dimension,
\(\left( u_{cg}(t), s_{cg}(t) \right)\) are the unspliced and spliced
expression functions given the change of time per cell and gene. We
restrict attention to piecewise-constant \(\alpha_{cg}(t)\) to capture
gene-specific activation and repression. We take special care to model a
gene- and cell-specific switching time that marks a single transition
from activation to repression by introducing a Bernoulli variable
\(k_{cg}\) to model unknown transcription state. We assume our
cell-by-gene data-matrix arrive as observations of Poisson-counts
related to the solution of the above ODEs at unknown times, which is
modeled as a relationship between an unknown latent time shared across
each cell, \(t_c\), and unknown gene-specific time-offsets \(t_{0,g}\)
where all read counts for a single cell occurred at an unknown, but
shared latent time \(t_c\). These relative times are also used to
parametrize the Bernoulli process for \(k_{cg}\). Importantly, we
recognize that the initial conditions are, in fact, unknown.

@@

Next we introduce the priors for the splicing parameters (where the
activation rate \(\alpha\) depends on the activation state \(k_{cg}\)
from above): \begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1.
  \end{cases}
\end{align}

@@

We define the ODE solution at time \(\tau_{cg}\) as: \begin{equation}
  \label{eq-ODE-solve}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

Next, we define the observation model that gives rise to the observed
counts as: \begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big)   \label{eq-u-hat-obs} \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big) \label{eq-s-hat-obs}
\end{align} Here, we use
\({u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}\) to denote the
observed unspliced and spliced counts for cell \(c\) and gene \(g\). We
use \(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}\) to
denote our generative model's prediction of these unspliced and spliced
expression levels. The generative process for modeling these observed
read counts given denoised gene transcript expression level
\(\hat{u}_{cg}, \hat{s}_{cg}\) considers the expected number of observed
reads for a given gene in a given cell as the number of transcripts
times the ratio of the cell's total reads to total transcripts.

The linear ODE in Eq. \ref{eq-rna} has an analytic solution under the
assumption that the transcription rate \(\alpha(t)\) is piece-wise
constant, and is derived by the authors of scVelo and a theoretical RNA
velocity study \citep{Bergen2020-pj, Li2021-qa}. For brevity, we present
the solution under constant transcription rate \(\alpha(t) = \alpha_0\),
and note that this can be combined to assemble the solution in the
piece-wise constant case. First, let \(C = u_0 - \frac{\alpha}{\beta}\).
Then, the solution to the ODE in Eq. \ref{eq-rna} is: \begin{equation}
\label{eq-rna-solution}
\begin{aligned}
  u(t) &= \frac{\alpha}{\beta} + C e^{-\beta t} \\
  s(t) &= \begin{cases}
    \frac{\alpha}{\gamma} + \frac{\beta}{\gamma - \beta} C e^{-\beta t} + \Big(s_0 - \frac{\alpha}{\gamma} - \frac{\beta}{\gamma - \beta} C\Big) e^{-\gamma t} & \text{if } \beta \neq \gamma \\
    \frac{\alpha}{\beta} + \Big(s_0 - \frac{\alpha}{\beta} + \beta C t\Big) e^{-\beta t} & \text{if } \beta = \gamma.
  \end{cases}
\end{aligned}
\end{equation} These solutions are the essential building block used to
compute all cell-gene-specific ODE solutions for the unspliced and
spliced expression levels in the generative model.

@@

Given observations
\(\tilde{X}_{cg} = \left( u_{cg}^{\text{(obs)}}, s_{cg}^{\text{(obs)}} \right)\)
over cells \(1 \dots C\) and genes \(1 \dots G\), we would like to
compute the posterior distribution over the random variables
\(\theta := \left( \theta_1, \ \dots, \theta_C \right)\) and
\(\psi := \left( \psi_1, \ \dots, \psi_G \right)\), where \begin{align}
\theta_c &= \left( t_{c}, \eta_{c}^{(u)}, \eta_{c}^{(s)} \right), \\
\phi_g &= \left( {t_{0,g}^{(0)}}, \Delta \text{switching}_{g}, 
          \alpha^{(0)}_{g}, \beta_{g}, \gamma_{g} \right),
\end{align} but exact Bayesian inference is intractable in this model.
We use Pyro to automatically integrate out the local discrete latent
variables \(k\), which is defined as the cell and gene transcriptional
state (see above), and approximate the posterior over the remaining
latent variables using variational inference
\citep{Bingham2018-id, Kucukelbir2016-bk}, which converts intractable
integrals into optimization problems that can be solved with
off-the-shelf tools. In variational inference, we maximize a tractable
bound (the evidence lower bound (ELBO)) on the Kullback-Leibler (KL)
divergence \[
\text{KL}\left( q_{\phi}(\theta, \psi) \mathrel{\Vert} 
  p(\theta, \psi \vert \tilde{X}_{cg}) \right)
\] between a parametric family of tractable probability distributions q
and the intractable true posterior \begin{align}
\varphi^* &= \operatorname{argmax}_{\varphi} \text{ELBO} \nonumber \\
&= \operatorname{argmax}_{\varphi} \bigg\{ E_{q_{\varphi}} 
     \bigg[ \log \left(p\left(\tilde{X}_{c g}, \theta, \psi \right) \right) \nonumber\\
&\qquad\qquad\qquad\qquad -\log \left(q_{\varphi}(\theta, \psi)\right) \bigg] \bigg\}
\end{align} We approximate our model's posterior distribution
\(p( \theta, \psi \vert
\tilde{X})\) with a tractable family of probability distributions
\(q_{\psi}(\theta, \phi) = q_{\phi_1}(\theta) q_{\phi_2}(\psi)\), where
\(q_{\phi_1}(\theta)\) is a product of independent univariate Gaussian
distributions with learnable location and scale parameters and
\(q_{\phi_2}(\psi)\) is a multivariate Gaussian distribution with a
learnable location parameter and a learnable low-rank covariance matrix.
We solve the resulting stochastic optimization problem using a version
of Automatic Differentiation Variational Inference (ADVI)
\citep{Kucukelbir2016-bk}: we obtain gradient estimates by
differentiating a Monte Carlo estimate of the ELBO and update the
variational parameters using stochastic gradient ascent. Our
implementation and experiments use generic variational families and
Monte Carlo ELBO and gradient estimators provided by Pyro
\citep{Bingham2018-id}, an open source library for probabilistic machine
learning, and a version built into Pyro of the adaptive stochastic
gradient ascent algorithm Adam augmented with gradient clipping to
enhance numerical stability during training.

@@

To benchmark Pyro -Velocity performance in predicting cell fate, we
generated posterior samples of quantitites of interest in the model. In
particular, we focus on posterior samples of \(i)\) spliced and
unspliced gene expression across all cells and genes, defined in Eq.
\ref{eq-ODE-solve} and denoted by
\(X_{cg}=\left(u_{cg}, s_{cg} \right)\) (used for phase portraits and
vector-field-based trajectory inference), \(ii)\) read counts, defined
in Eqs. \ref{eq-u-hat-obs} and \ref{eq-s-hat-obs} and denoted by
\(\hat{X}^{\text{(obs)}}_{cg}=\left(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg} \right)\)
(used for gene selection), and \(iii)\) RNA velocity, defined in Eq.
\ref{eq-ds-cg} as the time derivative of the spliced expression, denoted
by \(v_{cg}=\frac{ds_{cg}}{d t}(X_{cg})\). We generate 30 Monte Carlo
samples from the posterior predictive distribution
\(X, \hat{X}^{(\text{obs})}, v \sim p(X, \hat{X}^{(\text{obs})}, v \vert \theta, \psi) p(\theta, \psi) \approx q_{\phi}(\theta, \psi)\)
by sampling the posterior distribution of \(\theta\), \(\psi\) using the
trained model. We use these samples to compute statistics on the
posterior predictive distribution of our quantities of interest; these
statistics are used to evaluate the model's performance and uncertainty
in predicting cell fate.

@@

We prioritize the cell fate markers using two metrics: First, the
Pearson correlation between each gene's posterior mean of the denoised
spliced expression and posterior mean of the shared time; Second, the
negative mean absolute errors of each gene's observed spliced, and
unspliced read counts with posterior predictive samples of spliced and
unspliced raw read counts, i.e.
\(\mathcal{E}_g = \frac{1}{NC} \sum_{i=1}^N \sum_{c=1}^{C} \vert \tilde{X}_{cg} - \hat{X}_{cg}^{(\text{obs}), (i)} \vert\),
where we use \(N=30\) Monte Carlo samples. We first select the top
\(300\) genes with the highest negative mean absolute errors and then
rank the \(300\) genes based on the most positively correlated genes and
the least negatively correlated genes. We use the same strategy for
scVelo to rank the markers by the model likelihood to get the top 300
genes and then prioritize these genes by Pearson's correlation between
scVelo latent time and normalized expression.
